---
title: "CKME136 Household Debt Analysis - Initial Code/Results"
output: github_document
author: Daniel Christian 
---

# 1. Initial Dataset Preparation

I used three datasets from Statistics Canada and the BOC to explore the issue of rising household debt:

[Debt service indicators of households](www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1110006501)

[Financial indicators, quarterly, 1990 - 2012](www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610001001)

[Credit measures, Bank of Canada](www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1010011801)

## Debt Service Indicators
```{r}
debt = read.csv("Debt.csv")
debt1 = within(debt, rm(GEO, DGUID, UOM, UOM_ID, SCALAR_FACTOR, SCALAR_ID, VECTOR, COORDINATE, STATUS, SYMBOL, TERMINATED, DECIMALS))
colnames(debt1) = c("Date", "Adjustment", "Estimates", "Value")
unique(debt1$Estimates)
debt1$Estimates = factor(debt1$Estimates, levels = c("Household income", "Compensation of employees", "Net mixed income", "Property income received", "Current transfers received", "Household outlays", "Property income paid", "Current transfers paid", "Household final consumption expenditure", "Disposable income", "Plus: Interest paid", "Equals: disposable income before payment of interest", "Total debt payments", "Mortgage debt payments", "Non-mortgage debt payments", "Total obligated payments of principal", "Obligated mortgage principal payments", "Obligated non-mortgage principal payments", "Total interest paid", "Mortgage interest paid", "Non-mortgage interest paid", "Of which: consumer debt", "Debt service ratio", "Mortgage debt service ratio", "Non-mortgage debt service ratio", "Debt service ratio, interest only", "Mortgage debt service ratio, interest only", "Non-mortgage debt service ratio, interest only", "Of which: consumer debt service ratio, interest only"))
debt1$Estimates[10] = "Disposable income"
debt1$Estimates[22] = "Of which: consumer debt"
debt1$Date = as.Date(paste(debt1$Date, "-01", sep = ""))
View(debt1)

#filters to separate estimates by Adjustment, will be used later on. For the purposes of my inital results I will be focusing on data adjusted at annual rates. 
adfiltera = "Seasonally adjusted at annual rates"
adfilteru = "Unadjusted"

```




## Financial Indicators
```{r}
financial = read.csv("Financial.csv")
financial1 = within(financial, rm(GEO, DGUID, UOM, UOM_ID, SCALAR_FACTOR, SCALAR_ID, VECTOR, COORDINATE, STATUS, SYMBOL, TERMINATED, DECIMALS))
colnames(financial1) = c("Date", "Estimates", "Value")
unique(financial1$Estimates)
financial1$Estimates = factor(financial1$Estimates, levels = c("Debt to gross domestic product (GDP)", "Debt to personal disposable income", "Credit market debt to personal disposable income", "Consumer credit and mortgage liabilities to personal disposable income", "Net worth as a percentage of personal disposable income", "Debt to total assets", "Debt to net worth", "Credit market debt to net worth", "Consumer credit and mortgage liabilities to net worth", "Total assets to net worth", "Financial assets to net worth", "Financial assets to non-financial assets", "Owner's equity as a percentage of real estate", "Real estate as a percentage of personal disposable income"))
financial1$Date = as.Date(paste(financial1$Date, "-01", sep = ""))
View(financial1)

```




## Credit Measures
```{r}
Credit = read.csv("Credit.csv")
credit1 = within(Credit, rm(GEO, DGUID, UOM, UOM_ID, SCALAR_FACTOR, SCALAR_ID, VECTOR, COORDINATE, STATUS, SYMBOL, TERMINATED, DECIMALS))
credit1$Type.of.credit = factor(credit1$Type.of.credit, levels = c("Total business and household credit", "Household credit", "Residential mortgage credit", "Consumer credit", "Business credit", "Other business credit", "Short-term business credit"))
colnames(credit1) = c("Date", "Type of credit", "Adjustment", "Value")
unique(credit1$`Type of credit`)
credit1$Date = as.Date(paste(credit1$Date, "-01", sep = ""))
adcreditfiltera = "Seasonally adjusted"
#the adjustment column has differently phrased values warranting a different filter
credit1_a = credit1[credit1$Adjustment %in% adcreditfiltera,]
credit1_a = within(credit1_a, rm("Adjustment"))
View(credit1_a)

```
***





# 2. Consumer Debt Breakdown
                                  
Let's take a look at the different components of Canadian debt:

## A historical look at total consumer debt
```{r}
#Total Debt Payments of households from the Debt Service Indicators Dataset
tdfilter = "Total debt payments"
Total_debt_payments = debt1[debt1$Estimates %in% tdfilter,]
Total_debt_payments = within(Total_debt_payments, rm("Estimates"))
Total_debt_payments_a = Total_debt_payments[Total_debt_payments$Adjustment %in% adfiltera,]
Total_debt_payments_a = within(Total_debt_payments_a, rm("Adjustment"))
View(Total_debt_payments_a)
tdplot_a = plot(Total_debt_payments_a$Date, Total_debt_payments_a$Value, xlab = "Date", ylab = "Amount", main = "Total Debt Payment 1990-2020, Seasonally Adjusted", col = "red", type = "l")

```



```{r}
#Mortgage debt payments of households from the Debt Service Indicators Dataset
mdfilter = "Mortgage debt payments"
Mortgage_debt_payments = debt1[debt1$Estimates %in% mdfilter,]
Mortgage_debt_payments = within(Mortgage_debt_payments, rm("Estimates"))
Mortgage_debt_payments_a = Mortgage_debt_payments[Mortgage_debt_payments$Adjustment %in% adfiltera,]
Mortgage_debt_payments_a = within(Mortgage_debt_payments_a, rm("Adjustment"))
View(Mortgage_debt_payments_a)
mdplot_a = plot(Mortgage_debt_payments_a$Date, Mortgage_debt_payments_a$Value, xlab = "Date", ylab = "Amount", main = "Mortgage Debt Payment 1990-2020, Seasonally Adjusted", col = "red", type = "l")

```

```{r}
#Non-mortgage debt payments of households from the Debt Service Indicators Dataset
nmdfilter = "Non-mortgage debt payments"
nonmortgage_debt_payments = debt1[debt1$Estimates %in% nmdfilter,]
nonmortgage_debt_payments = within(nonmortgage_debt_payments, rm("Estimates"))
nonmortgage_debt_payments_a = nonmortgage_debt_payments[nonmortgage_debt_payments$Adjustment %in% adfiltera,]
nonmortgage_debt_payments_a = within(nonmortgage_debt_payments_a, rm("Adjustment"))
plot(nonmortgage_debt_payments_a$Date, nonmortgage_debt_payments_a$Value, xlab = "Date", ylab = "Amount", main = "Non-mortgage Debt 1990-2020, Seasonally Adjusted", col = "red", type = "l")

```

```{r}
#Putting all of the attributes together, compare
colnames(Total_debt_payments_a) = c("Date", "Total Debt Payments")
colnames(Mortgage_debt_payments_a) = c("Date", "Mortgage Debt Payments")
colnames(nonmortgage_debt_payments_a) = c("Date", "Non-mortgage Debt Payments")
Debt_breakdown = merge(x = Total_debt_payments_a, y = Mortgage_debt_payments_a, by = "Date", all.y = TRUE)
Debt_breakdown = merge(x = Debt_breakdown, y = nonmortgage_debt_payments_a, by = "Date", all.y = TRUE)
write.csv(Debt_breakdown, file = "Debt Breakdown.csv")
DebtB = read.csv("Debt Breakdown.csv")
DebtB = within(DebtB, rm("X"))
summary(DebtB)
plot(DebtB$Total.Debt.Payments, type = "l", ylim = c(20000, 250000), col = "red", xlab = "Date", ylab = "Amount", main = "Overall Debt Breakdown", xaxt = "n")
axis(1, at=1:length(DebtB$Date), labels = DebtB$Date)
lines(DebtB$Mortgage.Debt.Payments, col = "blue", type = "l")
lines(DebtB$Non.mortgage.Debt.Payments, col = "green", type = "l")
legend("topleft", legend = c("Total Debt Payments", "Mortgage Debt", "Non-mortgage Debt"), lty = 1, lwd = 2, pch = 21, col = c("red", "blue", "green"), ncol = 2, bty = "n", cex = 0.8, text.col = c("red", "blue", "green"), inset = 0.01)

```

From the data above it appears that over the past several decades credit debt has risen at a faster rate than mortgage debt, and so today makes up a larger portion of total household debt. Median levels of non-mortgage debt payments account for almost double those of mortgage debt payments. 
An interesting note here is that around 2008, both credit and mortgage debt hit a sudden peak due to the financial crisis (not only mortgage debt as much of the literature seems to focus on).




## Debt/ Net worth ratios over time
```{r}
#Total debt to net worth ratio from the Financial Indicators dataset
dnwfilter = "Debt to net worth"
debt_networth_ratio = financial1[financial1$Estimates %in% dnwfilter,]
debt_networth_ratio = within(debt_networth_ratio, rm("Estimates"))
colnames(debt_networth_ratio) = c("Date", "Total Debt/Networth Ratio")
plot(debt_networth_ratio$Date, debt_networth_ratio$`Total Debt/Networth Ratio`, xlab = "Date", ylab = "Ratio", main = "Total Debt to Net Worth Ratio, 1990-2012", type = "l", col = "red")

```

```{r}
#credit debt to net worth ratio from the Financial Indicators dataset
cdnwfilter = "Credit market debt to net worth"
credit_networth_ratio = financial1[financial1$Estimates %in% cdnwfilter,]
credit_networth_ratio = within(credit_networth_ratio, rm("Estimates"))
colnames(credit_networth_ratio) = c("Date", "Credit/Networth Ratio")
plot(credit_networth_ratio$Date, credit_networth_ratio$`Credit/Networth Ratio`, xlab = "Date", ylab = "Ratio", main = "Credit market debt to net worth Ratio, 1990-2012", type = "l", col = "red")

```

```{r}
#credit and mortgage debt to net worth ratio from the Financial Indicators dataset
cmnwfilter = "Consumer credit and mortgage liabilities to net worth"
creditm_networth_ratio = financial1[financial1$Estimates %in% cmnwfilter,]
creditm_networth_ratio = within(creditm_networth_ratio, rm("Estimates"))
colnames(creditm_networth_ratio) = c("Date", "Credit/mortgage to Net worth")
plot(debt_networth_ratio$Date, debt_networth_ratio$`Total Debt/Networth Ratio`, xlab = "Date", ylab = "Ratio", main = "Consumer Credit and Mortgage Liabilities to Net Worth, 1990-2012", type = "l", col = "red")

```

```{r}
#Putting all of the attributes together, compare
Networth_breakdown = merge(x = debt_networth_ratio, y = credit_networth_ratio, by = "Date", all.y = TRUE)
Networth_breakdown = merge(x = Networth_breakdown, y = creditm_networth_ratio, by = "Date", all.y = TRUE)
write.csv(Networth_breakdown, file = "Networth Breakdown.csv")
NetworthB = read.csv("Networth Breakdown.csv")
NetworthB = within(NetworthB, rm("X"))
summary(NetworthB)
plot(NetworthB$Total.Debt.Networth.Ratio, type = "l", ylim = c(15, 30), col = "red", xlab = "Date", ylab = "Ratio", main = "Net Worth Breakdown, 1990-2012", xaxt = "n")
axis(1, at=1:89, labels = NetworthB$Date)
lines(NetworthB$Credit.Networth.Ratio, col = "blue", type = "l")
lines(NetworthB$Credit.mortgage.to.Net.worth, col = "green", type = "l")
legend("topleft", legend = c("Total Debt/Networth Ratio", "Credit/Networth Ratio", "Credit+Mortgage/Networth Ratio"), lty = 1, lwd = 2, pch = 21, col = c("red", "blue", "green"), ncol = 2, bty = "n", cex = 0.8, text.col = c("red", "blue", "green"), inset = 0.01)
```

As the above data shows, debt to networth ratios are at their highest since the 1990s, with a median level of about 20%. Interestingly, mortgages make up a much more sizeable shift in the curve downwards once taken into the ratio consideration (the green curve).




## Historical debt service ratios 
The ratio of operating income available to debt servicing will be explored as well:
```{r}
#Historical debt service ratios from the Debt Service Indicators Dataset
dsrfilter = "Debt service ratio"
debt_service_ratio = debt1[debt1$Estimates %in% dsrfilter,]
debt_service_ratio = within(debt_service_ratio, rm("Estimates"))
debt_service_ratio_a = debt_service_ratio[debt_service_ratio$Adjustment %in% adfiltera,]
debt_service_ratio_a = within(debt_service_ratio_a, rm("Adjustment"))
colnames(debt_service_ratio_a) = c("Date", "Overall DSR")
plot(debt_service_ratio_a$Date, debt_service_ratio_a$`Overall DSR`, xlab = "Date", ylab = "Ratio", main = "Overall Debt Service Ratio, 1990-2020", type = "l", col = "red")

```

```{r}
#Historical Mortgage debt service ratios from the Debt Service Indicators Dataset
mdsrfilter = "Mortgage debt service ratio"
mortgage_dsr = debt1[debt1$Estimates %in% mdsrfilter,]
mortgage_dsr = within(mortgage_dsr, rm("Estimates"))
mortgage_dsr_a = mortgage_dsr[mortgage_dsr$Adjustment %in% adfiltera,]
mortgage_dsr_a = within(mortgage_dsr_a, rm("Adjustment"))
colnames(mortgage_dsr_a) = c("Date", "Mortgage DSR")
plot(mortgage_dsr_a$Date, mortgage_dsr_a$`Mortgage DSR`, xlab = "Date", ylab = "Ratio", main = "Mortgage Debt Service Ratio, 1990-2020", type = "l", col = "red")

```

```{r}
#Historical Non-mortgage debt service ratios from the Debt Service Indicators Dataset
nmdsrfilter = "Non-mortgage debt service ratio"
non_mortgage_dsr = debt1[debt1$Estimates %in% nmdsrfilter,]
non_mortgage_dsr = within(non_mortgage_dsr, rm("Estimates"))
non_mortgage_dsr_a = non_mortgage_dsr[non_mortgage_dsr$Adjustment %in% adfiltera,]
non_mortgage_dsr_a = within(non_mortgage_dsr_a, rm("Adjustment"))
colnames(non_mortgage_dsr_a) = c("Date", "Non-mortgage DSR")
plot(non_mortgage_dsr_a$Date, non_mortgage_dsr_a$`Non-mortgage DSR`, xlab = "Date", ylab = "Ratio", main = "Non-mortgage Debt Service Ratio, 1990-2020", type = "l", col = "red")

```

```{r}
#Putting all of the attributes together, compare
DSR_breakdown = merge(x = debt_service_ratio_a, y = mortgage_dsr_a, by = "Date", all.y = TRUE)
DSR_breakdown = merge(x = DSR_breakdown, y = non_mortgage_dsr_a, by = "Date", all.y = TRUE)
write.csv(DSR_breakdown, file = "DSR Breakdown.csv")
DSR = read.csv("DSR Breakdown.csv")
DSR = within(DSR, rm("X"))
summary(DSR)
plot(DSR_breakdown$`Overall DSR`, type = "l", ylim = c(0,15), col = "red", xlab = "Date", ylab = "% Change", main = "Debt Service Ratio 1990-2020, Seasonally Adjusted", xaxt = "n")
axis(1, at=1:119, labels = DSR_breakdown$Date)
lines(DSR_breakdown$`Mortgage DSR`, col = "blue", type = "l")
lines(DSR_breakdown$`Non-mortgage DSR`, col = "green", type = "l")
legend("bottomright", legend = c("Overall DSR", "Mortgage DSR", "Non-mortgage DSR"), lty = 1, lwd = 2, pch = 21, col = c("red", "blue", "green"), ncol = 2, bty = "n", cex = 0.8, text.col = c("red", "blue", "green"), inset = 0.01)

```





## Debt Breakdown Results

Overall consumer debt has been steadily growing, especially over the last decade where the rate of debt growth has increased. Also over the past decade non-mortgage debt has risen to higher levels than mortgage debt. According to historical debt service ratios, consumers over the past 2 decades have been allocating more of their income to servicing credit debt, while credit card debt represents an increasingly larger portion of their overall net worth.

After exploring the data it is clear that non-mortgage debt (specifically credit debt) is a growing concern over mortgage debt, and so will be the focus of my predictive model.
***





# 3. Credit Market Data
                                         
I acquired an additional statscan/BOC dataset "Credit Measures" in order to focus on consumer credit debt and use a larger pool of data for my predictice model. Looking at the set's attributes:
```{r}
unique(credit1_a$`Type of credit`)
```
There are two attributes in particular that are helpful here; [2]Household credit and [3]Consumer credit (where consumer credit is the measure of the individual agent and household credit is the aggregate measure of households)

## Exploring the attributes
```{r}
#Household credit
hcfilter = "Household credit"
household_credit = credit1_a[credit1_a$`Type of credit` %in% hcfilter,]
household_credit = within(household_credit, rm("Type of credit"))
colnames(household_credit) = c("Date", "Household Credit")
summary(household_credit)
plot(household_credit$Date, household_credit$`Household Credit`, type = "l", col = "red", xlab = "Date", ylab = "Amount", main = "Growth of Household Credit, 1969-2020")

```

```{r}
#Consumer credit
ccfilter = "Consumer credit"
consumer_credit = credit1_a[credit1_a$`Type of credit` %in% ccfilter,]
consumer_credit = within(consumer_credit, rm("Type of credit"))
colnames(consumer_credit) = c("Date", "Consumer Credit")
summary(consumer_credit)
plot(consumer_credit$Date, consumer_credit$`Consumer Credit`, type = "l", col = "red", xlab = "Date", ylab = "Amount", main = "Growth of Consumer Credit, 1969-2020")
```

```{r}
#Overall Credit debt 
Credit_breakdown = merge(x = household_credit, y = consumer_credit, by = "Date", all.y = TRUE)
write.csv(Credit_breakdown, file = "Credit Breakdown.csv")
CreditB = read.csv("Credit Breakdown.csv")
CreditB = within(CreditB, rm("X"))
summary(CreditB)
plot(CreditB$Household.Credit, type = "l", ylim = c(0,2500000), col = "red", xlab = "Date", ylab = "Amount", main = "Consumer Credit Breakdown 1969-2020", xaxt = "n")
axis(1, at=1:613, labels = CreditB$Date)
lines(CreditB$Consumer.Credit, col = "blue", type = "l")
legend("topleft", legend = c("Household Credit", "Consumer Credit"), lty = 1, lwd = 2, pch = 21, col = c("red", "blue"), ncol = 2, bty = "n", cex = 0.8, text.col = c("red", "blue"), inset = 0.01)
```
Household credit debt has been steadily rising with a rapid increase in the rate of debt growth following the 2008 financial crisis. 

Comparing median credit debt levels from 2005 to levels in 2019:
```{r}
hcredit_2005 = household_credit[(household_credit$Date > "2005-01-01" & household_credit$Date < "2005-12-01"),]
median(hcredit_2005$`Household Credit`)

```
```{r}
hcredit_2019 = household_credit[(household_credit$Date > "2019-01-01" & household_credit$Date < "2019-12-01"),]
median(hcredit_2019$`Household Credit`)
```
Median credit levels have risen by almost 250% in the span of 15 years.

                                         
                                         
                                         
                                         
# 4. Predictive Model

Next steps will be to use the caret package in R to make a predictive linear model of rising houshold credit levels.




